A molecular taxonomy of tumors independent of tissue-of-origin

### 2021-07-01 text/discussion.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the clusters are consistent with disease
progression. We were unable to identify any such trend in the number of
mutations, the relative staging or metastasis; and each cluster instead was
associated with unique combinations of pathways. Some cancer types are unevenly
distributed among the clusters, though we could not identify any cancers that
were exclusive to a single cluster. Only cancer types with the fewest samples
were found to be absent from one or more clusters at all.

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes
(_e.g. PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the
fact that they contribute to many cluster-specific pathways. This can only be
explained by less common genes complementing the unique combination of driver
genes in each tumor, and we speculate that many of these less common genes are
sensitive to increased mutation frequency in different tissues.

>>>>

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the clusters are consistent with disease
progression. We were unable to identify any such trend in the number of
mutations, the relative staging or metastasis; and each cluster instead was
associated with unique combinations of pathways. Some cancer types are unevenly
distributed among the clusters, though we could not identify any cancers that
were exclusive to a single cluster. Only cancer types with the fewest samples
were found to be absent from one or more clusters at all. Additionally, we annotated 
the clusters with patient attribute values (_e.g._ age of diagnosis, gender, and race) 
and were unable to identify any trends.

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes
(_e.g. PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the
fact that they contribute to many cluster-specific pathways. This can only be
explained by less commonly mutated genes complementing the unique combination of driver
genes in each tumor, and we speculate that many of these less commonly mutated genes are
sensitive to increased mutation frequency in different tissues. For example, the most 
frequently mutated genes (_TP53_, _PIK3CA_ and _MUC16_) in cluster 2 were also frequently 
mutated in several other clusters. Yet, pathway enrichment in cluster 2 was defined 
by less commonly mutated genes in the proteasomal degradation pathway (_e.g. PSMA3_, 
_PSMA1_ or _PSME4_). The ubiquitin proteasome system maintains cellular protein 
homeostasis by regulating protein turnover. Pathway mutations found in cluster 2 
(_e.g._ “B-catenin degradation”, “cellular response to hypoxia”, “regulation of cell cycle” 
and _etc_) covered several aspects of the cancer hallmarks described by Hanahan 
and Weinberg. These hallmarks included “sustaining proliferative signaling”, “evading 
tumor suppressors” and “resisting cell death” [@Bhattacharjee2014]. Cluster 
3 was defined by less commonly mutated genes related to ribosomal subunit. Pathway 
enrichment for cluster 3 included dysregulation in “rRNA processing in the nucleus 
and cytosol” and consequently, pathways involved in eukaryotic translation and 
nonsense mediated decay. Mutations in the ribosomal subunit have been associated 
with selective translation of mRNA to upregulate proliferation [@kampen2019]. 
Pathway enrichment in cluster 3 consist of hallmarks relating to “sustaining proliferative 
signaling”, “evading tumor suppressors”, “resisting cell death”, “deregulating 
cellular energetics”, and “genome instability and mutations”. Cluster 4 was defined 
by less commonly mutated genes related to nucleosome structure that resulted in 
pathway enrichment in DNA methylation and DNA double-strand break repair pathway. 
Cluster 4 was also enriched for “telomere maintenance”, a mechanism contributing to 
cancer progression by counteracting telomere shortening. Mutations in the 
nucleosome as found in cluster 4 have been proposed to hit all attributes of hallmarks of cancer. Cluster 
7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks: 
“sustaining proliferative signaling”, “evading tumor suppressors”, “resisting cell 
death”, and “inducing angiogenesis”. These clusters were strongly defined by 
mutations in _PI3K_ or _RAS_, respectively. Taken together, our clusters highlight 
the similarities and differences at a gene and pathway level between and within 
different cancer types – uncovering the molecular mechanism in cancer. In the future,
we hope to explore the defining features of the clusters by connecting the pathway 
mutations to cancer hallmarks.

### 2021-07-01 text/introduction.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
genes mutated at a greater rate than expected due to chance. It has been
estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Mutations in cancer driver
genes are commonplace in healthy tissue and correlate with age and environmental
exposures [@mutations-healthy-tissues].

@@

More recently, pan-cancer mutations and other data have been looked at as
disruptions of normal pathway activity. Sanchez-Vega et al. [@PCAWG-PATH] carried out exhaustive
analysis of pathway enrichment using whole genome data. Network approaches
address the sparseness of mutations, allowing genes to be influenced by
mutations in nearest network neighbors [@hotnet2;@hornNetSigNetworkbasedDiscovery2018;
@Iranzo-Network]. These methods are powerful especially when applied to smaller
numbers of tumors. One of the challenges in the field remains that there may be
thousands of moderate effect genes that occur at such low frequency that they
are impossible to detect using positive selection theory. Some researchers have
attempted to address this challenge by applying machine learning to cancer data
to discover groups of functionally related genes as they interact with larger
pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting fundamental disease phenotypes at the pathway level across cancers
with different origins at the cellular and tissue level.

>>>>

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
genes mutated at a greater rate than expected due to chance. Dees et al. [@deesMuSiC2012]
developed a pipline to separate driver mutations from passenger mutations. 
Tamborero et al. [@tamboreroOncodriveCLUST2013] expanded on this idea to identify 
tumor mutation biases towards specific regions of a protein sequence. Lawrence et 
al. [@lawrenceDiscoverySaturationAnalysis2014] focused on identifying cancer driver 
genes with intermediate mutation frequencies using tumor-normal pairs. It has been
estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Mutations in cancer driver
genes are commonplace in healthy tissue and correlate with age and environmental
exposures [@mutations-healthy-tissues].

@@

More recently, pan-cancer mutations and other data have been looked at as
disruptions of normal pathway activity. Pathway-centric analysis of tumors
provides additional benefits over gene-centric analysis: (1) It is significantly 
less noisy because it aggregates molecular events across multiple genes in the 
same pathway and (2) the identification of a potential casual mechanism is easier 
to interpret as genetic aberrations are linked to molecular pathways. In a perspective, 
Creixell et al. [@Creixell-pathwaynetworkanalysis] described approaches of pathway 
and network analysis applicable to next-generation sequencing data. The most common 
approach being fixed-gene enrichment analysis to identify over-represented pathways. 
Sanchez-Vega et al. [@PCAWG-PATH] carried out an exhaustive
analysis of pathway enrichment using whole genome data. Focusing on ten frequently 
altered canonical signaling pathways in cancer, they explored the mutation frequencies 
of coding regions of genes in these pathways for each cancer type. Network approaches
address the sparseness of mutations, allowing genes to be influenced by
mutations in nearest network neighbors [@hotnet2;@hornNetSigNetworkbasedDiscovery2018;
@Iranzo-Network]. These methods are powerful especially when applied to smaller
numbers of tumors. One of the challenges in the field remains that there may be
thousands of moderate effect genes that occur at such low frequency that they
are impossible to detect using positive selection theory. Some researchers have
attempted to address this challenge by applying machine learning to cancer data
to discover groups of functionally related genes as they interact with larger
pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting fundamental disease phenotypes at the pathway level across cancers
with different origins at the cellular and tissue level.

### 2021-07-01 text/manuscript.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

**Fig 5. Comparison of different cancers, simulated and actual results.**
Comparison of survival in four simulated cohorts. A) A group consisting of
randomly selected 30 year old women, unaffected (light blue) _vs._ cancer, 
($k = 2$, light orange), another group consisting of random 30-70 year-old
women, unaffecteds (dark blue) and cancer $k = 2$ (dark orange). B,C,D,E)
Compare random group with correct age distribution for the given cancer
(turqoise) to the model predicted survival for the given average k (blue) and
the actual survival for the patients in our dataset (orange). Differences with
actual data are likely due to a mixture of cluster identities which we exclude
from these simulations for simplicity. F) Effects are compared between cancer
type by showing the probability density for the posterior value of the $k_{tis}$
multiplier. Larger values correspond to decreased age-independent life
expectancy.

**Fig 6. Cluster specific $k_{cl}$ values for each cancer type.** The overall
$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific $k$
represents relative aggressiveness of each cluster within cancer type. Alpha
transparency (α) is set to reflect varying confidence in the posterior
distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N
≥ 25 α = 1.0.

>>>>

**Fig 5. Comparison of different cancers, simulated and actual results.**
Comparison of survival in six simulated cohorts. A) A group consisting of
randomly selected 30 year old women, unaffected (light blue) _vs._ cancer 
using the typical values from the posterior samples for a BRCA patient (yellow) 
and an artificial 10 years added to the effective age for a BRCA patient 
(dark blue), another group consisting of uniformly selected 30-70 year-old
women at random, unaffected (green) and cancer without (pink) and with (dark orange) 
the artificial 10 years. B) Posterior probability density of a 20 year old 
accelerated aging by cancer type (effective age). Red dashed line represents
age for unaffected individual, C) Posterior probability density of effective 
age rate for every additional year past 20 year old by cancer type. Red 
dashed line represents effective age rate for unaffected individual, D) 
Posterior probability density of the $k_{tis}$ multiplier. Values greater 
than 1 correspond to decreased age-independent life expectancy. Red dashed 
line represents $k_{tis}$ for unaffected individual. E,F,G,H) Comparison 
of our model to the actual survival for the patients in our dataset. We 
compare random group with correct age distribution for the given cancer 
(turqoise) to the model predicted survival for the given average effective 
age (blue) and the actual survival for the patients in our dataset (orange). 
Differences with actual data are likely due to a mixture of cluster 
identities which we exclude from these simulations for simplicity.

**Fig 6. Cluster specific $A_{cl}$ values for each cancer type.** The overall
effective age value is the product of $A_{0} x A_{cl} + A_{r} x (A^1 – 20)$. 
The cluster-specific A represents relative aggressiveness of each cluster within 
cancer type. Alpha transparency (α) is set to reflect varying confidence in the posterior
distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N
≥ 25 α = 1.0.

### 2021-07-01 text/survival.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

To illustrate the model, we simulate survival for four related groups (**Fig
5A**). We start with a cohort of women aged 30 randomly selected from the US
population, and show expected survival (light blue). We then use typical values
from the posterior samples to set the effective age of that cohort to the
effective age of a BRCA patient (yellow). Finally to illustrate how the model
varies by effective age we add an additional artificial 10 years to the BRCA
effective age (dark blue). Compare these results to a randomly selected group of
women ranging from 30 to 70 years old (green). Immediately, survival changes due
to the mixture of ages, without malignancy. By computing the effective ages for
this mixture of patients, we see that malignancy further reduces expected
lifespan (pink). Adding an additional 10 years to effective age would reduce the
lifespan even further (dark orange). Hence, simply due to the change in
distribution of age-at-diagnosis, we expect two equally deadly cancers to have
*different* survival curves. The model takes this into account and cancer
aggressiveness is estimated accurately without the confounding effect of age
through the effective age and aging rate parameters.

>>>>

To illustrate the model, we simulate survival for six related groups (**Fig
5A**). We start with a cohort of women aged 30 randomly selected from the US
population, and show expected survival (light blue). We then use typical values
from the posterior samples to set the effective age of that cohort to the
effective age of a BRCA patient (yellow). Finally to illustrate how the model
varies by effective age we add an additional artificial 10 years to the BRCA
effective age (dark blue). Compare these results to a uniformly selected group of
women ranging from 30 to 70 years old at random (green). Immediately, survival changes due
to the mixture of ages, without malignancy. By computing the effective ages for
this mixture of patients, we see that malignancy further reduces expected
lifespan (pink). Adding an additional 10 years to effective age would reduce the
lifespan even further (dark orange). Hence, simply due to the change in
distribution of age-at-diagnosis, we expect two equally deadly cancers to have
*different* survival curves. The model takes this into account and cancer
aggressiveness is estimated accurately without the confounding effect of age
through the effective age and aging rate parameters.
